\LaTeXe\ guide for the Journal of Functional Programming

### 2019-10-16 papers/jfp/7-experience.tex
Add 7.1 on Haskell as design language

Based on the theory developed in this paper we have extended \Shake to
become \Cloud \Shake, the first cloud-capable build system to support
both early cutoff and monadic dependencies~(\S\ref{sec-implementation-cloud}),
and used it to implement GHC's (very substantial) build system, Hadrian~\cite{hadrian}.
In this section we tie these papers together, reflecting on our
experience of turning theory into practice.

>>>>

Based on the theory developed in this paper we have extended \Shake to
become \Cloud \Shake, the first cloud-capable build system to support
both early cutoff and monadic dependencies~(\S\ref{sec-implementation-cloud}),
and used it to implement GHC's (very substantial) build system, Hadrian~\cite{hadrian}.
In this section we reflect on our experience of turning theory into practice.

\subsection{Haskell as a design language}\label{sec-design-lang}

Build systems are surprisingly tricky.  It is easy to waffle, and
remarkably hard to be precise. As this paper exemplifies, it is
possible to use Haskell as a \emph{design language}, to express quite
abstract ideas in a precise form -- indeed, precise enough to be executed.

Moreover, doing so is extremely beneficial.
The discipline of writing executable prototypes in
Haskell had a profound effect on our thinking. It forced misconceptions
to to surface early.  It required us to be explicit about side effects.
It gave us a huge incentive to design abstractions that had simple types
and an explicable purpose.

Consider, for example, our \hs{Task} type:
\begin{minted}[xleftmargin=10pt]{haskell}
data Task c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
\end{minted}
We started off with a much more concrete type, and explored
multiple variants.  During those iterations, this single line of
code gave us a tangible, concrete basis for productive conversations,
much more so than general conversations about ``tasks''.

It is also worth noting that we needed a rather expressive language,
to faithfully express the abstractions that seen natural in this
setting.  In the case of \hs{Task} we use a data constructor with a
polymorphic field; a higher-kinded type variable \hs{f :: * -> *}; an
even more abstracted kind for \hs{c :: (* -> *) -> Constraint}; and,
of course, type classes.

Moreover, when thinking about the type constructors over which \hs{f} might
usefully range, it turned out that we could adopt the \emph{existing} abstractions
of \hs{Functor}, \hs{Applicative}, \hs{Monad} and so on.
That in turn led us to think in a new way about the taxonomy
of build systems -- see \S~\ref{sec-why-polymorphism}.

The effect of using a concrete design language went well beyond
merely \emph{expressing} our ideas: it \emph{directly influenced}
our thinking.  For example, here are definitions for a scheduler and rebuilder,
from \S~\ref{sec-concrete-impl}:
\begin{minted}[fontsize=\small,xleftmargin=5pt]{haskell}
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task@\,@(MonadState ir) k v
\end{minted}
These powerful and modular abstractions, which ultimately formed part
of the conceptual structure of the paper, emerged fairly late in the
project, as we repeatedly reviewed, redesigned, and refactored our
executable prototype.  It is hard to believe that we could have
developed them without the support of Haskell as a design language.
